Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-23 text/ms.tex
Edit and write introduction and discussion

Black swans are unexpected extreme events with dramatic consequences
\citep{taleb2007,sornette2009}. One of the most striking black swans in ecology
was the asteroid marking the mass extinction of the K-T boundary (REF). Today,
it is the extremeness of climate, in concert with shifts in mean temperature,
that is expected to cause the greatest societal damage \citep{ipcc2012}.
However, whereas extremes in the physical environment such as wave height, rain
fall, and temperature are present and widely accepted
\citep{gaines1993,katz2005}, it remains unclear the extent to which ecological
systems buffer or suffer from black swans \citep{nunez2012}.

%Existing evidence for black swans or heavy tails in population dynamics is
%mostly anecdotal or for particular systems.
There is compelling anecdotal evidence for ecological black swans, and
quantitative evidence for a limited number of systems, but systematic evidence
has been elusive. \citet{doak2008} reviewed a collection of field experiment
population dynamic surprises and conclude that these are more common than we
typically assume. \citet{nunez2012} compiled a list of major events on geologic
and contemporary scales that could be considered ecological black swans. Other
work has identified heavy-tail dynamics for particular systems. For example,
\citet{segura2013} identified heavy tails in time series of marine microbe
populations and \citet{drake2014} identified heavy tails in time to extinction
for experimental waterflea populations. However, systematic evidence has been
limited. \citet{keitt1998} tested for power-law tail distributions in breeding
bird abundance, but later worked showed the apparent heavy tails were a result
of mixing populations of different species \citep{allen2001}. A later large
scale analysis of the Global Population Dynamics Database (GPDD) by
\citet{halley2002} found little evidence for heavy-tail distributions in longer
abundance time series.

@@

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
population dynamics model fitting and risk forecasting assumes that population
dynamics are not heavy tailed (e.g.~REF, REF, REF). Alternatively, dynamics
might be heavy tailed, but a paucity of statistical methods for the frequently
available short time series has limited our ability to detect them. For
example, a commonly applied probability distribution to detect the shape of
time series tails (the generalized extreme value distribution) requires lumping
data into the most extreme event per unit time (e.g.~heaviest rainfall per
year) and therefore requires a considerable quantity of data (REF). Previous
large-scale work on the distribution of abundance trends focused on model
comparison to identify parsimonious descriptions, but has not quantified the
magnitude of heavy tails on a continuous scale (REF). Furthermore, information
criteria are poor at distinguishing subtle differences in population dynamics
for the relatively short time series that are most readily available (REF
CHANGE THIS?).
%bring in Ward - great method but data intensive, can't be applied to most
%individual time series from the GPDD
%plankton --- hideous distribution to fit, even harder in population dynamics
%framework -and what about seasonal cycles, density dependence?
%GPDD - model competition and finding most parsimonious explanation from set of
%hypotheses --- but only \emph{very} heavy from not heavy
%GEV --- super data intensive
%tipping point and regime shift methods --- say something\ldots{} not
%appropriate for most of the data available

@@

We assembled the documented causes of black swan population dynamics in seven
populations with high probabilities of heavy tails (Table 1). The majority of
documented events were downward black swans and the majority involved
a confluence of events. For example, a shortage of nest sites for the shag
long-necked bird in the UK was thought to reduce population productivity and
make the population vulnerable when a red-tide hit in 1968 \citep{potts1980}.
The population experienced a black swan downswing, but the new availability of
nest sites resulted in a rapid population upswing \citep{potts1980}. As
another example, an interaction between environmentally and predation-mediated
population cycles are thought to have caused a downward black swan for a water
vole population \citep{saucy1994}. Other black swans were the result of
a sequence of extreme climate events on their own. For example, severe winters
in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings
in grey heron abundance in the UK \citep{stafford1971}. Our analysis finds
that the last event was a combination of two black swan events in a row and it
took the population three times longer to recover than predicted
\citep{stafford1971}.

@@

Our results clarify, but do not disagree with previous related analyses.
Previous work with an older version of the GPDD assessed the distribution of
abundance timeseries but focused on identifying if the lognormal distribution
was the most frequently parsimonious model \citep{halley2002}. For heavy-tailed
distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy
distribution and the four-parameter Levy stable distribution and found little
information criteria support for these distributions in longer time series.
However, black swan events are by definition rare, and the majority of time
series in the GPDD are short for these purposes. Therefore, we would not expect
to observe black swans in a large proportion of populations. Furthermore, by
probabilistically quantifying process noise tail behaviour on a continuous
spectrum, and allowing for non-stationary time series and density dependence,
our analysis allows for a more nuanced description of the evidence of
ecological black swans. In an earlier study, \citet{keitt1998} described heavy
(power law) tails in breeding bird population abundance. However, this finding
was challenged by \citet{allen2001}, who showed that mixing data across species
could create falsely generate heavy tails. This rebuttal supports our use of
the t distribution to represent heavy tails, since a t distribution can be
represented by a mixture of normal distributions \citep[with
inverse-gamma-distributed variances,][]{gelman2014}.

We might expect to observe black swan dynamics in ecological time series
because of unmodelled intrinsic properties of populations or extrinsic forces
acting on populations. Since a t-distribution can be formed by a mixture of
normal distributions, we could observe heavy tails if we miss some underlying
mixture of intrinsic processes \citep{allen2001}. That process might be an
aggregation of populations across space, or population diversity, or some
intrinsic change in population variability through time. Extrinsic forces
could also cause black swan dynamics \citep[e.g.][]{nunez2012}. These forces
could be extreme themselves. For example, extreme climate, predation from or
competition with other species experiencing black swans, or sharp changes in
human pressure such hunting, fishing, or habitat destruction might cause black
swans. Alternatively, the interaction of multiple ``normal'' extrinsic
forces could give rise to black swan ecological dynamics. This could occur if
the interaction nature is synergistic \citep[e.g.][]{kirby2009} or even if
non-synergistic interactions experience a rare alignment \citep{denny2009}.

There are a number of caveats when considering the generality of our results.
First, the GPDD data represent a taxonomically and geographically biased sample
of populations --- the longer time series we focus on are dominated by
commercially and recreationally important species and a disproportionate number
of populations are located in the United Kingdom. Although we expect that we
would find strong but rare evidence for black swans in other large taxonomic or
geographic samples, the common forces driving those black swans (e.g.~severe
winters in Table~1) likely differ. Second, in a large database with disparate
data sources, we cannot ensure the quality of every data point. Recording
mistakes undoubtedly comprise some of the black swans we detected. Conversely,
recording mistakes could mask heavy tails if initial observations are assumed
to be recording errors and are discarded or altered. Third, the temporal scale
of observation and population dynamics vary considerably across populations in
the GPDD and this likely influences the detection of heavy tails. For example,
if we make frequent observations relative to generation time (e.g.~for many
large-bodied mammals) we will average across generations and perhaps miss
black swans. Furthermore, if we census populations infrequently relative to
generation time (e.g.~many insects in the GPDD) the recorded data may average
across extreme and less-extreme events and dampen black swan dynamics.

Our results suggest important future issues to address about ecological black
swans. Given that black swans can and do occur, can we forecast the probability
of black swans in space and time? Furthermore, what management policies allow
us to detect them quickly after they happen? Secondly, can we isolate the
components of ecological dynamics that experience black swans events by moving
from phenomenological models such as the Gompertz to mechanistic models, that
take into account, for example, recruitment? We expect that greater insight
into the mechanisms and covariates of ecological black swans may be best
obtained through specific geographic and taxonomic subsets of data where longer
time series are available \citep[e.g.][]{segura2013}. Finally, what is the
impact of allowing for black swans in forecasts of ecological risk? Financial
analysts are concerned with the shape of financial-return distribution tails
because these impact estimates of risk --- the probability of a specific
magnitude of undesired event occurring \citep{rachev2008}. It is possible that
a comparable focus in ecology would alter our perception of ecological
forecasts such as extinction risk.

Policy implications

unlikely to forecast black swans, at least with any useful advanced warning
but we might be able to determine probabilities and develop policy that is
robust to, encourages rapid detection, and that is flexible enough to respond
quickly to black swans

example of earth quakes - spatial planning, early warning systems, education
systems, a general idea about the probability in given places, build buildings
that are robust

Suggests that it's important to have policy that is robust to the uncertainty
about black swans and that can respond and adapt rapidly following black swans

foot and mouth disease and earthquakes?

>>>>

Black swans are unexpected extreme events with dramatic consequences
\citep{taleb2007,sornette2009}. One of the most striking black swans in ecology
was the asteroid marking the mass extinction of the K-T boundary (REF). Today,
it is the extremeness of climate, in concert with shifts in mean temperature,
that is expected to cause the greatest societal damage \citep{ipcc2012}.
But, while extremes in the physical environment such as wave height, rain
fall, and temperature are present and widely accepted
\citep{gaines1993,katz2005}, it remains unclear the extent to which ecological
systems buffer or suffer from black swans \citep{nunez2012}.

%Existing evidence for black swans or heavy tails in population dynamics is
%mostly anecdotal or for particular systems.
There is compelling anecdotal evidence for ecological black swans, but
systematic evidence has been elusive. \citet{doak2008} reviewed a collection of
field experiment population dynamic surprises and concluded that surprises are
more common than we typically assume. \citet{nunez2012} compiled a list of
major events on geologic and contemporary scales that could be considered
ecological black swans. Other work has identified heavy-tail dynamics for
particular systems. For example, \citet{segura2013} identified heavy tails in
time series of marine microbe populations and \citet{drake2014} identified
heavy tails in time to extinction for experimental waterflea populations.
However, systematic evidence is limited. \citet{keitt1998} tested for power-law
tail distributions in breeding bird abundance, but later work showed the
apparent heavy tails were a result of mixing populations of different species
\citep{allen2001}. A subsequent large scale analysis of the Global Population
Dynamics Database (GPDD) by \citet{halley2002} found little evidence for
heavy-tail distributions in longer abundance time series.

@@

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
model fitting and risk forecasting assumes that population
dynamics are not heavy tailed (e.g.~REF, REF, REF).
Alternatively, dynamics might be heavy tailed, but a paucity of statistical
methods appropriate for short time series has limited our
ability to detect them.
For example, (1) the generalized extreme value (GEV) distribution has been
applied to environmental data (REF).
The GEV describes the distribution of the most extreme
event per time interval (e.g.~heaviest rainfall per year), but this approach
requires long time series that can be condensed into time intervals (REF).
(2) Heavy- and thin-tailed distributions have been competed with each
other (REF), but information criteria likely have
trouble accurately distinguishing such subtle differences in population
dynamics for short time series (REF).
(3) A catastrophic mixture distribution model allows one to elegantly quantify
the probability that population events are catastrophic is also data intensive
\citep{ward2007}.
(4) Other research has focussed on tipping points and regime shifts, but such
methods are usually focussed on entire communities or ecosystems (e.g.\ REF).
Furthermore, black swans are more general than tipping points --- they don't
necessarily imply a permanent change of state.

% For example, a commonly applied probability distribution to detect the shape of
% time series tails (the generalized extreme value distribution) requires lumping
% data into the most extreme event per unit time (e.g.~heaviest rainfall per
% year) and therefore requires a considerable quantity of data (REF). Previous
% large-scale work on the distribution of abundance trends focused on model
% comparison to identify parsimonious descriptions, but has not quantified the
% magnitude of heavy tails on a continuous scale (REF). Furthermore, information
% criteria are poor at distinguishing subtle differences in population dynamics
% for the relatively short time series that are most readily available (REF
% CHANGE THIS?).
%bring in Ward - great method but data intensive, can't be applied to most
%individual time series from the GPDD plankton --- hideous distribution to fit,
%even harder in population dynamics framework -and what about seasonal cycles,
%density dependence? GPDD - model competition and finding most parsimonious
%explanation from set of hypotheses --- but only \emph{very} heavy from not
%heavy GEV --- super data intensive tipping point and regime shift methods ---
%say something\ldots{} not appropriate for most of the data available

@@

We assembled the documented causes of black swan population dynamics in seven
populations with high probabilities of heavy tails (Table 1). The majority of
documented events were downward black swans and the majority involved
a confluence of events. For example, a shortage of nest sites for the shag
long-necked bird in the UK was thought to reduce population productivity and
make the population vulnerable when a red tide hit in 1968 \citep{potts1980}.
The population experienced a black swan downswing, but the new availability of
nest sites resulted in a rapid population upswing \citep{potts1980}. As
another example, an interaction between environmentally and predation-mediated
population cycles are thought to have caused a downward black swan for a water
vole population \citep{saucy1994}. Other black swans were the result of
a sequence of extreme climate events on their own. For example, severe winters
in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings
in grey heron abundance in the UK \citep{stafford1971}. Our analysis finds
that the last event was a combination of two black swan events in a row and it
took the population three times longer to recover than predicted
\citep{stafford1971}.

@@

Our results clarify previous related analyses. Previous work with an older
version of the GPDD assessed the distribution of abundance time series but
focused on identifying if the lognormal distribution was the most frequently
parsimonious model \citep{halley2002}. For heavy-tailed distributions,
\citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the
four-parameter Levy stable distribution and found little information criteria
support for these distributions in longer time series. However, black swan
events are by definition rare, and the majority of time series in the GPDD are
short for these purposes. Therefore, we would not expect to observe black swans
in a large proportion of populations. Furthermore, by probabilistically
quantifying process noise tail behaviour on a continuous spectrum, and allowing
for non-stationary time series and density dependence, our analysis allows for
a more nuanced description of the evidence of ecological black swans. In an
earlier study, \citet{keitt1998} described heavy (power law) tails in breeding
bird population abundance. However, this finding was challenged by
\citet{allen2001}, who showed that mixing data across species could create
falsely generate heavy tails. This rebuttal supports our use of the
t distribution to represent heavy tails, since a t distribution can be
represented by a mixture of normal distributions \citep[with
inverse-gamma-distributed variances,][]{gelman2014}.

We might expect to observe black swan dynamics in ecological time series
because of either unmodelled intrinsic properties of populations or extrinsic
forces acting on populations. Since a t-distribution can be formed by a mixture
of normal distributions \citep{gelman2014}, we could observe heavy tails if we
miss some underlying mixture of intrinsic processes \citep{allen2001}. That
process might be an aggregation of populations across space, or population
diversity, or some intrinsic change in population variability through time.
Extrinsic forces could also cause black swan dynamics
\citep[e.g.][]{nunez2012}. These forces could be extreme themselves. For
example, extreme climate, predation from or competition with other species
experiencing black swans, or sharp changes in human pressure such hunting,
fishing, or habitat destruction might cause black swans. Alternatively, the
interaction of multiple ``normal'' extrinsic forces could give rise to black
swan ecological dynamics. This could occur if the interaction nature is
synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions
experience a rare alignment \citep{denny2009}.

There are a number of caveats when considering the generality of our results.
First, the GPDD data represent a taxonomically and geographically biased sample
of populations --- the longer time series we focus on are dominated by
commercially and recreationally important species and a disproportionate number
of populations are located in the United Kingdom. Although we expect that we
would find strong but rare evidence for black swans in other large taxonomic or
geographic samples, the common forces driving those black swans (e.g.~severe
winters in Table~1) likely differ. Second, in a large database with disparate
data sources, we cannot ensure the quality of every data point. Recording
mistakes undoubtedly comprise some of the black swans we detected. Conversely,
recording mistakes could mask heavy tails if initial observations are assumed
to be recording errors and are discarded or altered. Third, the temporal scales
of observation and population dynamics vary considerably across populations in
the GPDD and this likely influences the detection of heavy tails. For example,
if we make frequent observations relative to generation time (e.g.~for many
large-bodied mammals) we will average across generations and perhaps miss
black swans. Furthermore, if we census populations infrequently relative to
generation time (e.g.~many insects in the GPDD) the recorded data may average
across extreme and less-extreme events and dampen black swan dynamics.

Our results suggest important future issues to address about ecological black
swans. Given that black swans do occur, can we forecast the probability of
black swans in space and time? Furthermore, what management policies allow us
to detect them quickly after they happen? Secondly, can we isolate the
components of ecological dynamics that experience black swans events by moving
from phenomenological models such as the Gompertz to mechanistic models, that
take into account, for example, recruitment? We expect that greater insight
into the mechanisms and covariates of ecological black swans may be best
obtained through specific geographic and taxonomic subsets of data where longer
time series are available \citep[e.g.][]{segura2013}. Finally, what is the
impact of allowing for black swans in forecasts of ecological risk? Financial
analysts are concerned with the shape of financial-return distribution tails
because these impact estimates of risk --- the probability of a specific
magnitude of undesired event occurring \citep{rachev2008}. It is possible that
a comparable focus in ecology would alter our perception of ecological
forecasts such as extinction risk.

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines focussed on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans. A similar focus could benefit resource management once our ability to
predict the spatial probability and covariates of ecological black swans
improves. Second, we can choose management policy that is robust to heavy
tails. For example, setting target population abundances that are appropriately
far from critical limits may buffer black swan events (REF), and
maintaining genetic, phenotypic, and behavioural diversity may allow some
components of populations persist when others are affected by extreme disease
or environmental forces (REF). Finally, extreme and `unexpected', `surprising', or
`counterintuitive' ecological dynamics offer a tremendous opportunity to learn
about ecological systems, evaluate when our models break down, and tune future
management policy (REF, REF). In recent decades ecology has moved toward
focussing on aspects of variance in addition to mean responses (REF). Our
results and other recent work suggests that an added focus on extremes
represents the next frontier, particularly in the face expected increases in
extreme climate (REF).

%ecological resource management can draw from other disciplines focussed on
%heavy tails, design policy that is robust to heavy tails, and take black swan
%events as opportunities for insight.
%The 'portfolio effect' is often used in the context of dampening variability
%(REFs), but the ultimate portfolio effect is maintaining some stability in the
%face of black swan events.
%unlikely to forecast black swans, at least with any useful advanced warning
%but we might be able to determine probabilities and develop policy that is
%robust to, encourages rapid detection, and that is flexible enough to respond
%quickly to black swans
%example of earth quakes - spatial planning, early warning systems, education
%systems, a general idea about the probability in given places, build buildings
%that are robust
%Suggests that it's important to have policy that is robust to the uncertainty
%about black swans and that can respond and adapt rapidly following black swans
%foot and mouth disease and earthquakes?
